The large-scale structure of the Universe in dispersion measure space

### 2015-06-03 writeup/dm_space_distortions.tex
Cafe edits.

    The recently discovered fast radio bursts are a new class of radio
    transient that, due to their high dispersion from intervening plasma,
    have been interpreted as
    originating from cosmological distances.  The rate of fast radio bursts
    has been estimated to be several thousand over the whole sky per day.  If
    cosmological, the sources of these events should trace the large-scale
    structure of the
    Universe. Furthermore, because the dispersion measure (DM) can be
    used as a proxy for radial distance (like redshift),
    clustering can be studied in three dimensions. Due to inhomogeneities in
    the plasma's density, dispersion measure is an imperfect proxy for radial
    distance and we show that this leads to additional clustering terms---the
    dispersion-measure-space distortions. We calculate the DM-space
    power spectra for a toy model of electron and FRB clustering and show that
    the clustering signal could be detected in a survey of $10^4$ FRB
    events, as is expected to be produced by the Canadian Hydrogen Intensity
    Mapping Experiment.

@@

However, redshift is not the only measure of radial distance. In particular,
dispersion of electromagnetic signals due to intervening plasma
is readily measured from fast radio bursts (FRB) which are a class
of radio transient first reported by \citet{2007Sci...318..777L}.
There have since been 
10 such bursts reported
\citep{2012MNRAS.425L..71K,2013Sci...341...53T,
2014ApJ...790..101S, 2014ApJ...792...19B, 
2015MNRAS.447..246P,2015ApJ...799L...5R} and total rate of bursts
been estimated to be $3.3^{+5}_{-2.5} \times 10^3$
per sky per day \cite{2015arXiv150500834R}. 
The dispersion measure
(DM) is a line-of-site integral of the free electron density and can thus be
used as a distance estimate, as is routinely done with galactic pulsars.
FRB events are observed to have dispersion measures of order
$1000\,{\rm pc} / {\rm cm}^3$, greatly in excess of
the galactic expectation of order $100\,{\rm pc} / {\rm cm}^3$, depending on
galactic latitude
\cite{2014arXiv1412.4829D}.
They must therefore be of extra-galactic origin, with their
dispersion produced either by the
inter-galactic medium (IGM) \citep{2013Sci...341...53T} or the
environment of the source
\citep{2014ApJ...785L..26L,2015arXiv150101341P,2015arXiv150505535C}.
If the dispersion is dominated by the inter-galactic medium,
then the average DM--distance relation can be modelled
\citep{2014ApJ...797...71Z,2004MNRAS.348..999I, 2003ApJ...598L..79I}
and the sources are at
cosmological distances of a few gigaparsecs. This would permit
the study of the clustering of FRB sources in three dimensions which should trace the
large-scale structure of the Universe on linear scales.

@@

\emph{Dispersion measure space} is the three dimensional coordinates,
$\vec x_s$ inferred from the dispersion measure assuming that the electrons are
homogeneous. This only affects the radial coordinate such that $\vec x_s = \hat
n \chi_s$ with $\chi_s$ defined by the equation
\begin{equation}
    {\rm DM}(\chi_s) = \int_0^\chi \ud\chi' a(\chi')^2 \bar{n}_e(\chi').
\end{equation}

@@

Here we consider the sensitivity of a survey with a total of $10^4$ dispersion
measures from FRB events distributed uniformly over half the sky. This is in rough
approximation to what could be achieved by the Canadian Hydrogen Intensity
Mapping Experiment's now funded FRB instrument over a three year survey.
To give an idea of the overall sensitivity of the survey, we collapse
$C^{ss}_{\ell,ij}$ to a single function of $\ell$ by taking a weighted average
of all pairs of radial bins. We use bin widths of $100\,{\rm Mpc}/h$ over the
range $500\,{\rm Mpc}/h$ to $3500\,{\rm Mpc}/h$.
For weights in the radial bin average, we use the signal over noise-squared
at $\ell=100$, which maximizes the signal to noise at that $\ell$ and
is near optimal at other multipoles.
We exclude the $i = j$ bin pairs
as they would be contaminated by noise bias in a real survey.
This effectively discards the contribution from the local term. The result is shown
in Figure~\ref{f:sensitivity}.

@@

We would like to thank Ue-Li Pen and Mathew McQuinn for helpful discussions.
%
KWM is supported by the Canadian Institute for Advanced Research, Global Scalars
Program.

>>>>

    The recently discovered fast radio bursts (FBRs) are a new class of radio
    transient that, due to their high dispersion from intervening plasma,
    have been interpreted as
    originating from cosmological distances.  The rate of fast radio bursts
    has been estimated to be several thousand over the whole sky per day.  If
    cosmological, the sources of these events should trace the large-scale
    structure of the
    Universe.
    We show here that, even without redshifts, these bursts can be used to
    study the three-dimensional cluster of matter in the Universe.
    This is because the dispersion of radio waves as they travel through the
    Universe, like redshift, can be
    used as a proxy for radial distance. 
    Due to inhomogeneities in
    the plasma's density, dispersion is an imperfect proxy for radial
    distance and we show that this leads to calculable dispersion-space
    distortions in the apparent clustering. We calculate the dispersion-space
    power spectra for a simple model of electron and FRB clustering and show that
    the clustering signal could be detected in a survey of $10^4$ FRB
    events. This rate is \ldots as is expected to be produced by the Canadian Hydrogen Intensity
    Mapping Experiment.

@@

However, redshift is not the only measure of radial distance. In particular,
dispersion of electromagnetic signals due to intervening plasma
is readily measured from fast radio bursts (FRB) which are a class
of radio transient first reported by \citet{2007Sci...318..777L}.
There have since been 
10 such bursts reported
\citep{2012MNRAS.425L..71K,2013Sci...341...53T,
2014ApJ...790..101S, 2014ApJ...792...19B, 
2015MNRAS.447..246P,2015ApJ...799L...5R} and total rate of bursts
been estimated to be $3.3^{+5}_{-2.5} \times 10^3$
per sky per day \cite{2015arXiv150500834R}. 
The dispersion measure
(DM) is a line-of-site integral of the free electron density and can thus be
used as a distance estimate, as is routinely done with galactic pulsars.
FRB events are observed to have dispersion measures of order
$1000\,{\rm pc} / {\rm cm}^3$, greatly in excess of
the galactic expectation of order $100\,{\rm pc} / {\rm cm}^3$, depending on
galactic latitude
\cite{2014arXiv1412.4829D}.
They must therefore be of extra-galactic origin, with their
dispersion produced either by the
inter-galactic medium (IGM) \citep{2013Sci...341...53T} or the
environment of the source
\citep{2014ApJ...785L..26L,2015arXiv150101341P,2015arXiv150505535C}.
If the dispersion is dominated by the inter-galactic medium,
then the average DM--distance relation can be modelled
\citep{2014ApJ...797...71Z,2004MNRAS.348..999I, 2003ApJ...598L..79I}
and the sources are at
cosmological distances of a few gigaparsecs. This would permit
the study of the clustering of FRB sources in three dimensions which should trace the
large-scale structure of the Universe on linear scales. Note that although we
frame our discussions in terms of FRBs, dispersion may be

@@

\emph{Dispersion measure space} is the three dimensional coordinates,
$\vec x_s$ inferred from the dispersion measure assuming that the electrons are
homogeneous. This only affects the radial coordinate such that $\vec x_s = \hat
n \chi_s$ with $\chi_s$ defined by the equation
\begin{equation}
    {\rm DM}(\chi_s) = \int_0^{\chi_s} \ud\chi' a(\chi')^2 \bar{n}_e(\chi').
\end{equation}

@@

We consider the sensitivity of a survey with a total of $10^4$ dispersion
measures from FRB events distributed uniformly over half the sky. This is in rough
approximation to what could be achieved by the Canadian Hydrogen Intensity
Mapping Experiment's now funded FRB instrument over a three year survey.
To give an idea of the overall sensitivity of the survey, we collapse
$C^{ss}_{\ell,ij}$ to a single function of $\ell$ by taking a weighted average
of all pairs of radial bins. We use bin widths of $100\,{\rm Mpc}/h$ over the
range $500\,{\rm Mpc}/h$ to $3500\,{\rm Mpc}/h$.
For weights in the radial bin average, we use the signal over noise-squared
at $\ell=100$, which maximizes the signal to noise at that $\ell$ and
is near optimal at other multipoles.
We exclude the $i = j$ bin pairs
as they would be contaminated by noise bias in a real survey.
This effectively discards the contribution from the local term. The result is shown
in Figure~\ref{f:sensitivity}.

@@

We thank Ue-Li Pen and Mathew McQuinn for helpful discussions.
%
KWM is supported by the Canadian Institute for Advanced Research, Global Scalars
Program.
